---
title: "Inference for numerical data"
output:
  pdf_document: default
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
---

## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=TRUE}
load("more/nc.RData")
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

1.  What are the cases in this data set? How many cases are there in our sample?  

*Exercise 1 Answer:*  
  
Each case is the information recorded about a birth during 2004 in North Carolina.  There are 1,000 cases in the sample.  

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary, eval=TRUE}
summary(nc)
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?  
  
*Exercise 2 Answer:*  
  
```{r bp, eval=TRUE}  
boxplot(weight~habit,data=nc, main = "Weight by Smoking Status of Mother", ylab = "Weight in Lbs.")
```  
  
The medians, min, and interquartile ranges seem fairly similar, but the max is considerably higher for nonsmoking mothers. There are some high-side outliers for nonsmoking mothers. There are also way more data points for nonsmoking mothers (873 observations) than for smoking mothers (126 observations).  

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

```{r by-means, eval=TRUE}
by(nc$weight, nc$habit, mean)
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test
.

## Inference

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length`.  

*Exercise 3 Answer:*  
First we check if the observations are independent. The data is from a simple random sample and consists of less than 10% of all cases (presumably there were 10,000 or more births in North Carolina in 2004). Next, we make sure both distributions are normal. See histograms and notes below:  
  
```{r ex3, eval=TRUE}
# Subset nonsmoker and smoker observations
nc_non <- subset(nc, habit == "nonsmoker")
nc_smk <- subset(nc, habit == "smoker")
# Create histograms
hist(nc_non$weight)
hist(nc_smk$weight)
```  
  
Though both histograms exhibit some skew, the sample sizes are reasonably large enough that statistical inference analyses can be pursued.  


4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.  
  
*Exercise 4 Answer:*  

*H~0~*:    $\mu_{diff} = 0$    There is no difference in average birth weights for children born from mothers who did or did not smoke (null hypothesis)  

*H~A~*:    $\mu_{diff} \neq 0$    There is a difference in average birth weights for children born from mothers who did or did not smoke (alternative hypothesis)  
  
This is s a two-tailed test, because the research question is asking if there is a difference in either direction.  
  

Next, we introduce a new function, `inference`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r inf-weight-habit-ht, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

Let's pause for a moment to go through the arguments of this custom function. 
The first argument is `y`, which is the response variable that we are 
interested in: `nc$weight`. The second argument is the explanatory variable, 
`x`, which is the variable that splits the data into two groups, smokers and 
non-smokers: `nc$habit`. The third argument, `est`, is the parameter we're 
interested in: `"mean"` (other options are `"median"`, or `"proportion"`.) Next
we decide on the `type` of inference we want: a hypothesis test (`"ht"`) or a 
confidence interval (`"ci"`). When performing a hypothesis test, we also need 
to supply the `null` value, which in this case is `0`, since the null 
hypothesis sets the two population means equal to each other. The `alternative`
hypothesis can be `"less"`, `"greater"`, or `"twosided"`. Lastly, the `method` 
of inference can be `"theoretical"` or `"simulation"` based.

5.  Change the `type` argument to `"ci"` to construct and record a confidence 
interval for the difference between the weights of babies born to smoking and 
non-smoking mothers.  
  
*Exercise 5 Answer:*  
  
```{r inf-weight-habit-ci, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical")
```  
  

By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$)
. We can easily change this order by using the `order` argument:

```{r inf-weight-habit-ci-ord, eval=TRUE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical", 
          order = c("smoker","nonsmoker"))
```

* * *

## On your own

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. Note that since you're doing inference 
on a single population parameter, there is no explanatory variable, so you can 
omit the `x` variable from the function.

-   Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level by adding a new argument 
to the function: `conflevel = 0.90`.

-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

-   Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `inference` 
function, report the statistical results, and also provide an explanation in 
plain language.  
  
  
*Number 1 Answer:*  
  
```{r oyo1, eval=TRUE, tidy=FALSE}
inference(y = nc$weeks, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical")
```
  
Interpretation:  We can be 95% confident that the pregnancy length in weeks for a mother is between 38.1528 and 38.5165.  
  

*Number 2 Answer:*  
  
```{r oyo2, eval=TRUE, tidy=FALSE}
inference(y = nc$weeks, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical", conflevel = 0.90)
```
  
  
*Number 3 Answer:*  
  
```{r oyo3, eval=TRUE, tidy=FALSE}
inference(y = nc$gained, x = nc$mature, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```  
  
Because the p-value exceeds the $\alpha$ value, we fail to reject the null hypothesis. The data does not provide strong evidence that the weight gain during pregnancy differed for younger mothers vs. mature mothers.  
  
*Number 4 Answer:*  
  
```{r oyo4, eval=TRUE, tidy=FALSE}
# Create data frame containing observations with younger moms
nc_younger <- subset(nc, mature == "younger mom")
# Create data frame of obsverations with younger moms
nc_mature <- subset(nc, mature == "mature mom")
# Run summary statistics for mage for both data frames
summary(nc_younger$mage)
summary(nc_mature$mage)
# Max age for younger moms is 34, and min age for mature moms is 35, so the 
# age cut-off for younger and mature mothers is 35. If the mother is 35 or 
# older, she is classified as mature, otherwise she is classified as younger
```  
  
*Number 5 Answer:*  
  
```{r oyo5, eval=TRUE, tidy=FALSE}
inference(y = nc$visits, x = nc$lowbirthweight, est = "mean", type = "ht", null = 0, alternative = "twosided", method = "theoretical")
```  
  
The inference test indicates that the number of hospital visits during the pregnancy was significantly lower for mothers who gave birth to low weight babies than the number of hospital visits during the pregnancy for mothers who gave birth to babies which were not low weight.  
  



<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for OpenIntro by Mine &Ccedil;etinkaya-Rundel from a lab 
written by the faculty and TAs of UCLA Statistics.
</div>
